### 2016-09-15 Chapters/QuantumMechanics.tex
Polish

Molecular quantum mechanics is concerned with the microscopic motion of
nuclei and electrons in molecules and is thus an intrinsically many-body
theory. The problem is clearly very complicated to solve, as the number of
interactions to be considered is large.
Since the nuclei are much heavier than the electrons, the \gls{BO} or
clamped-nuclei approximation is typically assumed. The nuclei are fixed
in a specific configuration (called a \emph{molecular geometry}) and
treated as static electric sources.\autocite{Born1927-ce}
This separation of motions results in an electron-only Hamiltonian and
an electronic wave function that depends \emph{parametrically} on the
nuclear positions. The eigenvalue of this Hamiltonian, also dependent on
the nuclear positions, is the \gls{PES} of the system, that is the
potential function in which the nuclei move.\autocite{Szabo1989-vl,
McWeeny1992-oj}
In first quantization, the general expression for the clamped-nuclei,
$N$-electron, molecular electronic Hamiltonian is:
\begin{alignat}{2}\label{eq:fq-hamiltonian}
  H = \sum_{i=1}^N h(\vect{r}_i)
  + \frac{1}{2}\sum_{i\neq j}^N g(\vect{r}_i, \vect{r}_j)
  + V_\mathrm{NN},
  \quad&
  V_\mathrm{NN} =
  \frac{1}{2}\sum_{A \neq B}^{N_\mathrm{nuclei}}\frac{Z_A Z_B}{|\vect{R}_A - \vect{R}_B|}
\end{alignat}
The last term is the nuclear repulsion energy, a constant shift of the
energy scale in the \acrshort{BO} approximation.
As noted by \citeauthor{Saue2011-qg}, the same general expression for
the clamped-nuclei Hamiltonian Eq.~\eqref{eq:fq-hamiltonian} is valid in
the nonrelativistic and relativistic no-pair
regimes.\autocite{Sucher1980-vf, Saue2011-qg}
In the nonrelativistic domain one defines the one- and two-electron
operators as:\autocite{Szabo1989-vl, McWeeny1992-oj, Konishi2009-zb}
\begin{subequations}
 \begin{align}
   h(\vect{r}_i) &=
   -\frac{1}{2}\nabla^2_i + V_\mathrm{Ne}(\vect{r}_i)
   = -\frac{1}{2}\nabla^2_i - \sum_{A=1}^{N_\mathrm{nuclei}}\frac{Z_A}{|\vect{R}_A -
   \vect{r}_i|} \label{eq:h} \\
   g(\vect{r}_i, \vect{r}_j) &= \frac{1}{|\vect{r}_i - \vect{r}_j|},
   \label{eq:g-coulomb}
 \end{align}
\end{subequations}
while in the relativistic regime with the \gls{DC}
approximation:\autocite{Dyall2007-tu, Konishi2009-zb, Reiher2014-cp}
\begin{subequations}
 \begin{align}
   h(\vect{r}_i) &= (\mat{\beta} - I_4)\si{\electronmass}c^2 +
   c(\mat{\alpha}_i\cdot\vect{p}_i) + V_{\mathrm{Ne}}(\vect{r}_i)I_4 \\
   g(\vect{r}_i, \vect{r}_j) &= \frac{I_4\cdot I_4}{|\vect{r}_i - \vect{r}_j|}
   \label{eq:g-r-coulomb}
 \end{align}
\end{subequations}
The Dirac matrices, expressed in terms of the vector $\mat{\sigma}$
collecting the three Pauli spin matrices, were introduced:
\begin{alignat}{2}
  \mat{\alpha}_i =
  \begin{pmatrix}
    0_2 & \mat{\sigma}_i \\
    \mat{\sigma}_i & 0_2
  \end{pmatrix},
  \quad&
  \mat{\beta} =
  \begin{pmatrix}
    I_2 & 0_2 \\
    0_2 & -I_2
  \end{pmatrix},
\end{alignat}
and the operators are implicitly assumed to embed the proper no-pair,
positive-energy manifold projectors.\autocite{Sucher1980-vf}
Since these are fermionic Hamiltonians, the spin-statistics theorem
dictates the corresponding eigenfunctions to be
antisymmetric.
Using the expansion theorem,\autocite{Gross1991-hi} we seek the wave
function for the $N$-electron system as a linear combination of
$N$-electron \emph{Slater determinants}.

It the language of second quantization, we assume that a, not
necessarily complete, set of orthonormal one-particle states is
available. These are usually called \glspl{MO}.
In general, we would like to embed as many of the symmetries of the
system into the one-particle basis. This is accomplished by requiring
the one-particle space to be the common eigenbasis for a complete set of
suitable, commuting one-particle operators.\autocite{Gross1991-hi}
Any $N$-particle state can be constructed by distributing particles
into one-particle states and accordingly labelling them by the number of
particles per one-particle states present.
This is the so-called \gls{ON} vector representation:
\begin{equation}
  \ket{\vect{n}} = \ket{n1,n2,\ldots},
\end{equation}
and the span of all \acrshort{ON} vectors is called the \emph{Fock
space} $F^N$, a space isomorphic to the Hilbert space $H^N$ of
$N$-particle wave functions.
To preserve consistency, the \emph{vacuum} state with no particles:
\begin{equation}
 \ket{\mathrm{vac}} = \ket{0,0,\ldots}
\end{equation}
is included in the construction of Fock space.
Any $N$-particle state can be generated by application of
\emph{creation} operators $\cons{i}$. Creation operators increase the
occupation number in the one-particle state $i$ by $1$, up to a phase.
Conversely, \emph{annihilation} operators $\anni{i}$, defined as the Hermitian
conjugates of the creation operators, decrease occupation in state $i$
by $1$, again up to a phase. Clearly, annihilation of any one-particle
state in the vacuum results yields $0$.
Pauli's exclusion principle is enforced by the \emph{canonical
anticommutation relations}:
\begin{subequations}
  \begin{align}
    \cons{r}\cons{s} + \cons{s}\cons{r} &= 0 \\
    \anni{r}\anni{s} + \anni{s}\anni{r} &= 0 \\
    \cons{r}\anni{s} + \anni{s}\cons{r} &= \delta_{rs},
  \end{align}
\end{subequations}
showing how the construction of the Fock space representation
embeds antisymmetry at the operator level and not in the wave functions.
Finally, the second-quantized molecular electronic Hamiltonian in the \acrlong{BO}
approximation is:\autocite{Helgaker2000-tz, Shavitt2009-mr}
\begin{equation}\label{eq:sq-hamiltonian}
  H = \sum_{rs}h_{rs}\cons{r}\anni{s}
  + \frac{1}{2}\sum_{rstu}g_{rstu}\cons{r}\cons{t}\anni{u}\anni{s}
  + V_\mathrm{NN}
\end{equation}
where the matrix elements are given as integrals over the chosen
one-particle basis:
\begin{subequations}
 \begin{align}
   h_{rs} &= \int \diff\vect{r} \phi_r^\dagger(\vect{r}) h(\vect{r})\phi_s(\vect{r}) \label{eq:one-el}\\
   g_{rstu} &=
   \int\diff\vect{r}\int\diff\vect{r}^\prime
   \Omega_{rs}(\vect{r}) g(\vect{r},\vect{r}^\prime) \Omega_{tu}(\vect{r}^\prime) \label{eq:two-el} \\
  \Omega_{rs}(\vect{r}) &= \phi_r^\dagger(\vect{r})\phi_s(\vect{r}). \label{eq:overlap}
 \end{align}
\end{subequations}
The second-quantized Hamiltonian Eq.~\eqref{eq:sq-hamiltonian} is thus a
\emph{projected} operator, exact only within the subspace determined by
the chosen one-particle set.~\autocite{Gross1991-hi, Helgaker2000-tz}

@@

In the \acrlong*{CC} method we construct our correlated wave function as
an exponential mapping on top of a single reference
function:\autocite{Crawford2000-ey, Helgaker2000-tz, Bartlett2007-mz,
Shavitt2009-mr}
\begin{equation}
  \ket{\mathrm{CC}} = \expo{T}\ket{\mathrm{HF}},
\end{equation}
out of convenience we assume, here and in the following, that the
reference function is a closed-shell \acrshort{HF} wave function.
The \emph{cluster} operator appearing in the exponential is given as:
\begin{equation}
  T = \sum_{p=1}^\mathcal{M} T_p = \sum_{p=1}^\mathcal{M}\sum_{\mu_p}\tamp{\mu_p}\cluster{\mu_p}
\end{equation}
that is, as a truncated sum of excitation operators, $\cluster{\mu_p}$,
times the corresponding cluster amplitudes, $\tamp{\mu_p}$.
Here $\mu_p$ is the $\mu$-th excitation at the $p$-th excitation level and
$\mathcal{M}$ is the truncation level.
One of the main strengths of the \acrshort{CC} model is its size-extensivity
which stems directly from the exponential nature of the wave
operator.~\autocite{Shavitt2009-mr, Helgaker2000-tz}

>>>>

Molecular quantum mechanics is concerned with the microscopic motion of
nuclei and electrons in molecules and is thus an intrinsically many-body
theory. The problem is clearly very complicated to solve, as the number of
interactions to be considered is large.
Since the nuclei are much heavier than the electrons, the \gls{BO} or
clamped-nuclei approximation is typically assumed. The nuclei are fixed
in a specific configuration (called a \emph{molecular geometry}) and
treated as static electric sources.\autocite{Born1927-ce}
This separation of motions results in an electron-only Hamiltonian and
an electronic wave function that depends \emph{parametrically} on the
nuclear positions. The eigenvalue of this Hamiltonian, also dependent on
the nuclear positions, is the \gls{PES} of the system, that is the
potential function in which the nuclei move.\autocite{Szabo1989-vl,
McWeeny1992-oj}
In first quantization, the general expression for the clamped-nuclei,
$N$-electron, molecular electronic Hamiltonian is:
\begin{alignat}{2}\label{eq:fq-hamiltonian}
  H = \sum_{i=1}^N h(\vect{r}_i)
  + \frac{1}{2}\sum_{i\neq j}^N g(\vect{r}_i, \vect{r}_j)
  + V_\mathrm{NN},
  \quad&
  V_\mathrm{NN} =
  \frac{1}{2}\sum_{A \neq B}^{N_\mathrm{nuclei}}\frac{Z_A Z_B}{|\vect{R}_A - \vect{R}_B|}
\end{alignat}
The last term is the nuclear repulsion energy, a constant shift of the
energy scale in the \acrshort{BO} approximation.
As noted by \citeauthor{Saue2011-qg}, the same general expression for
the clamped-nuclei Hamiltonian Eq. \eqref{eq:fq-hamiltonian} is valid in
the nonrelativistic and relativistic no-pair
regimes.\autocite{Sucher1980-vf, Saue2011-qg}
In the nonrelativistic domain one defines the one- and two-electron
operators as:\autocite{Szabo1989-vl, McWeeny1992-oj, Konishi2009-zb}
\begin{subequations}
 \begin{align}
   h(\vect{r}_i) &=
   -\frac{1}{2}\nabla^2_i + V_\mathrm{Ne}(\vect{r}_i)
   = -\frac{1}{2}\nabla^2_i - \sum_{A=1}^{N_\mathrm{nuclei}}\frac{Z_A}{|\vect{R}_A -
   \vect{r}_i|} \label{eq:h} \\
   g(\vect{r}_i, \vect{r}_j) &= \frac{1}{|\vect{r}_i - \vect{r}_j|},
   \label{eq:g-coulomb}
 \end{align}
\end{subequations}
while in the relativistic regime with the \gls{DC}
approximation:\autocite{Dyall2007-tu, Konishi2009-zb, Reiher2014-cp}
\begin{subequations}
 \begin{align}
   h(\vect{r}_i) &= (\mat{\beta} - I_4)\si{\electronmass}c^2 +
   c(\mat{\alpha}_i\cdot\vect{p}_i) + V_{\mathrm{Ne}}(\vect{r}_i)I_4 \\
   g(\vect{r}_i, \vect{r}_j) &= \frac{I_4\cdot I_4}{|\vect{r}_i - \vect{r}_j|}
   \label{eq:g-r-coulomb}
 \end{align}
\end{subequations}
The Dirac matrices, expressed in terms of the vector $\mat{\sigma}$
collecting the three Pauli spin matrices, were introduced:
\begin{alignat}{2}
  \mat{\alpha}_i =
  \begin{pmatrix}
    0_2 & \mat{\sigma}_i \\
    \mat{\sigma}_i & 0_2
  \end{pmatrix},
  \quad&
  \mat{\beta} =
  \begin{pmatrix}
    I_2 & 0_2 \\
    0_2 & -I_2
  \end{pmatrix},
\end{alignat}
and the operators are implicitly assumed to embed the proper no-pair,
positive-energy manifold projectors.\autocite{Sucher1980-vf}
Since these are fermionic Hamiltonians, the spin-statistics theorem
dictates the corresponding eigenfunctions to be
antisymmetric.
Using the expansion theorem,\autocite{Gross1991-hi} we seek the wave
function for the $N$-electron system as a linear combination of
$N$-electron \emph{Slater determinants}.

It the language of second quantization, we assume that a, not
necessarily complete, set of orthonormal one-particle states is
available. These are usually called \glspl{MO}.
In general, we would like to embed as many of the symmetries of the
system into the one-particle basis. This is accomplished by requiring
the one-particle space to be the common eigenbasis for a complete set of
suitable, commuting one-particle operators.\autocite{Gross1991-hi}
Any $N$-particle state can be constructed by distributing particles
into one-particle states and accordingly labelling them by the number of
particles per one-particle states present.
This is the so-called \gls{ON} vector representation:
\begin{equation}
  \ket{\vect{n}} = \ket{n1,n2,\ldots},
\end{equation}
and the span of all \acrshort{ON} vectors is called the \emph{Fock
space} $F^N$, a space isomorphic to the Hilbert space $H^N$ of
$N$-particle wave functions.
To preserve consistency, the \emph{vacuum} state with no particles:
\begin{equation}
 \ket{\mathrm{vac}} = \ket{0,0,\ldots}
\end{equation}
is included in the construction of Fock space.
Any $N$-particle state can be generated by application of
\emph{creation} operators $\cons{i}$. Creation operators increase the
occupation number in the one-particle state $i$ by $1$, up to a phase.
Conversely, \emph{annihilation} operators $\anni{i}$, defined as the Hermitian
conjugates of the creation operators, decrease occupation in state $i$
by $1$, again up to a phase. Clearly, annihilation of any one-particle
state in the vacuum results yields $0$.
Pauli's exclusion principle is enforced by the \emph{canonical
anticommutation relations}:
\begin{subequations}
  \begin{align}
    \cons{r}\cons{s} + \cons{s}\cons{r} &= 0 \\
    \anni{r}\anni{s} + \anni{s}\anni{r} &= 0 \\
    \cons{r}\anni{s} + \anni{s}\cons{r} &= \delta_{rs},
  \end{align}
\end{subequations}
showing how the construction of the Fock space representation
embeds antisymmetry at the operator level and not in the wave functions.
Finally, the second-quantized molecular electronic Hamiltonian in the \acrlong{BO}
approximation is:\autocite{Helgaker2000-tz, Shavitt2009-mr}
\begin{equation}\label{eq:sq-hamiltonian}
  H = \sum_{rs}h_{rs}\cons{r}\anni{s}
  + \frac{1}{2}\sum_{rstu}g_{rstu}\cons{r}\cons{t}\anni{u}\anni{s}
  + V_\mathrm{NN}
\end{equation}
where the matrix elements are given as integrals over the chosen
one-particle basis:
\begin{subequations}
 \begin{align}
   h_{rs} &= \int \diff\vect{r} \phi_r^\dagger(\vect{r}) h(\vect{r})\phi_s(\vect{r}) \label{eq:one-el}\\
   g_{rstu} &=
   \int\diff\vect{r}\int\diff\vect{r}^\prime
   \Omega_{rs}(\vect{r}) g(\vect{r},\vect{r}^\prime) \Omega_{tu}(\vect{r}^\prime) \label{eq:two-el} \\
  \Omega_{rs}(\vect{r}) &= \phi_r^\dagger(\vect{r})\phi_s(\vect{r}). \label{eq:overlap}
 \end{align}
\end{subequations}
The second-quantized Hamiltonian Eq. \eqref{eq:sq-hamiltonian} is thus a
\emph{projected} operator, exact only within the subspace determined by
the chosen one-particle set.\autocite{Gross1991-hi, Helgaker2000-tz}

@@

In the \acrlong*{CC} method we construct our correlated wave function as
an exponential mapping on top of a single reference
function:\autocite{Crawford2000-ey, Helgaker2000-tz, Bartlett2007-mz,
Shavitt2009-mr}
\begin{equation}
  \ket{\mathrm{CC}} = \expo{T}\ket{\mathrm{HF}},
\end{equation}
out of convenience we assume, here and in the following, that the
reference function is a closed-shell \acrshort{HF} wave function.
The \emph{cluster} operator appearing in the exponential is given as:
\begin{equation}
  T = \sum_{u=1}^\mathcal{M} T_u = \sum_{u=1}^\mathcal{M}\sum_{\mu_u}\tamp{\mu_u}\cluster{\mu_u}
\end{equation}
that is, as a truncated sum of excitation operators, $\cluster{\mu_p}$,
times the corresponding cluster amplitudes, $\tamp{\mu_p}$.
Here $\mu_p$ is the $\mu$-th excitation at the $p$-th excitation level and
$\mathcal{M}$ is the truncation level.
One of the main strengths of the \acrshort{CC} model is its size-extensivity
which stems directly from the exponential nature of the wave
operator.\autocite{Shavitt2009-mr, Helgaker2000-tz}
